Recurrent Deep Learning Models and its applications

### 2025-03-29 paper2/main.tex
paper update

% Literature Review
\subsection{Literature Review}
% Backpropagation Through Time
\subsubsection{Backpropagation Through Time}
BPTT is one of the most important algorithms used for training RNNs. Dating back to the original effort to expand the typical backpropagation algorithm, BPTT has been formulated to handle the difficulties of temporal sequences that are inherent in sequential data \parencite{werbos1990bptt}. This algorithm allows RNNs in learning sequence dependent data by unfold the network over time steps and then updating weights matrix through the gradient of loss function with respect to the variable \parencite{rumelhart1986backpropagation}.
\\[2ex]
\textbf{Conceptual Framework of BPTT}
\\[1ex]
BPTT works based on the technique of treating an RNN as a deep feedforward network for across multiple time steps. In the forward pass, the RNN, like other artificial neuronal network, applies operation over the data input in sequence, bringing changes in its own state variables at every time step, depending on the input and the previous state of its general working state or hidden state. This sequential processing produces outputs and stores the internal states of the network in any period \parencite{werbos1990bptt}.

This unfolds the RNN to construct a traditional Feedforward Neural Network where we can apply backpropagation through time. Below is the conceptual idea of BPTT in RNN.
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\textwidth]{../Pic/pic1.png} % Replace "example-image" with your image file
    \caption{Unfolded RNN}
\end{figure}

\begin{longtable}{|c|c|c|}
    \hline
    \textbf{Notation} & \textbf{Meaning} & \textbf{Dimension}\\
    \hline
    $U$          & Weight matrix for input to hidden state       & $input\ size\times hidden\ unites$\\
    $W$          & Weight matrix for hidden to hidden state      & $hidden\ units\times hidden\ unites$\\
    $V$          & Weight matrix for hidden state to output state& $hidden\ units\times number\ of\ class$\\
    $x_t$        & Input vector at time t                        & $input\ size\times 1$\\
    $h_t$        & Hidden state output at time t                 & $hidden\ units\times 1$\\
    $b_h$        & Bias term for hidden state                    & $hidden\ units\times 1$\\
    $b_y$        & Bias term for output state                    & $number\ of\ class\times 1$\\
    $\hat{o}_y$  & Output at time t                              & $number\ of\ class\times 1$\\
    $\hat{y}_t$  & Output at time t                              & $hidden\ units\times 1$\\
    $\mathcal{L}$& Loss at time t                                & $scalar$\\
    \hline
    \caption{Unfolded RNN}
\end{longtable}
% Forward pass
\noindent \textbf{Forward Pass}
\\[1ex]
During the forward pass, the RNN processes the input sequence sequentially, computing hidden states and output at each timestep:
\begin{equation}
    h_t = f(U^Tx_{t}+W^Th_{t-1}+b_h)
\end{equation}
\begin{equation}
    \hat{y}_t = f(V^Th_t+b_y)
\end{equation}
\newline  % Computing the loss function
\noindent \textbf{Computing the loss function}
\\[1ex]
Assuming the loss is computed only at the final timestep t:
\begin{equation}
    \mathcal{L}_t = L(y_t, \hat{y}_t)
\end{equation}
In order to do backpropagation through time to tune the parameters in RNN, we need to calculate the partial derivative of loss function $\mathcal{L}$ with respect to the differently parameters.\\
\newline  % Backward pass using the chain rule
\noindent \textbf{Backward pass using the chain rule}
\\[1ex]
Using the chain rule for computing the gradient.\\
Partial derivative of loss function $\mathcal{L}$ with respect to $W$ (hidden to hidden state) at time 2. 
\begin{equation}
    \dfrac{\partial\mathcal{L}_2}{\partial W} = \dfrac{\partial\mathcal{L}_2}{\partial y_2} \cdot \dfrac{\partial y_2}{\partial h_2} \cdot \dfrac{\partial h_2}{\partial h_1} \cdot \dfrac{\partial h_1}{\partial W}
\end{equation}
By mathematic induction 
\begin{equation}
    \dfrac{\partial\mathcal{L}_t}{\partial W} = \dfrac{\partial\mathcal{L}_t}{\partial y_t} \cdot \dfrac{\partial y_t}{\partial h_t} \cdot (\sum_{i=1}^{t} \dfrac{\partial h_t}{\partial h_i} \cdot \dfrac{\partial h_i}{\partial W})
\end{equation}
Where 
\begin{equation}
    \dfrac{\partial h_t}{\partial h_i} = \prod_{j=i+1}^{t} \dfrac{\partial h_j}{\partial h_{j-1}}
\end{equation}

@@

@@

\subsubsection{RNN encoder decoder model}
\begin{figure}[!htb]
    \centering
    \includegraphics[width=1\textwidth]{../Pic/RNN_encoder_decoder.jpg} % Replace "example-image" with your image file
\end{figure}
RNN Encoder-Decoderis a RNN based architecture which commonly used for handle sequence to sequence problem, for example machine translation, summarization, and question answering. The designed of this architecture is good for handle the input sequence and the output sequence is not the same length, leveraging the sequential nature of data. 
\\[1ex]
In this architecture, there are two main components, the first component is rnn encoder, the usage of the encoder is to encode the input sequence into a fixed length vector. And the decoder is to decode the vector into the target sequence. 
\\[1ex]
\textbf{Encoder}
\\[1ex]
The encoder is a RNN based block, which processes the input sequence step by step and calculate the hidden state representation for each input token. Let the input sequence be $X = [x_1, x_2, \cdots, x_T]$ where $T$ is the length of the input sequence, and $x_t$ is the input at time step $t$. We use the below formula to computes the hidden states $h_t$.
\\[1ex]
For the Vanilla RNN encoder
\begin{equation}
    h_t = f_{enc}(x_t, h_{t-1})
\end{equation}

where the $f_{enc}$ is the recurrent transition function, (e.g. LSTM, GRU or Vanilla RNN cell), and $h_t$ is the hidden state at time step $t$. And then the final hidden state of the encoder, $h_T$ will act as the context vector.
\begin{equation}
    c = h_T
\end{equation}
\textbf{Decoder}
\\[1ex]
The decoder is the second component of RNN encoder decoder, the usage of the decoder is to generates the output sequence step by step, conditioned on the context vector $c$ and its previous hidden state. 
\begin{equation}
    s_t = f_{dec}(y_{t-1}, s_{t-1}, c)
\end{equation}
where $f_{dec}$ is the recurrent function of the decoder, and it is just like the encoder part, $y_{t-1}$ is the embedding of the previous output token, $s_{t-1}$ is the decoder's hidden state. And $c$ is the context vector from the encdoer, which remains constant throughout decoding. 
\\[1ex]
After the model computed the $s_t$, the model will compute the output probability distribution.
\begin{equation}
    P(y_t|y_1, c) = softmax(W_os_t+b_o)
\end{equation}
And the predicted token $y_t = arg maxP(y_t)$, the process will repeat until the decoder predicts the end of sequence token $(<EOS>)$.

@@

\textbf{Bahdanau (Additive) Attention}
\\[1ex]
% This method proposed by \parencite{bahdanau2016neuralmachinetranslationjointly}, in which this attention uses a feedforward network to compute the attention scores. 
\\[1ex]
\begin{equation}
    e_{t',t} = v_a^Ttanh(W_as_{t'-1}+U_ah_t)
\end{equation}                    
\textbf{Luong (Multiplicative) Attention}
\\[1ex]
% This method is introduced by Luong et al \parencite{luong2015effectiveapproachesattentionbasedneural}. This method uses the dot product to compute the attention scores.
\begin{equation}
    e_{t',t} = s_{t'-1}^Th_t
\end{equation}
\textbf{Scaled Dot-Product Attention}
\\[1ex]

>>>>

% Literature Review
\subsection{Literature Review}
% Backpropagation Through Time
\subsubsection{Backpropagation Through Time}
BPTT is one of the most important algorithms used for training RNNs. Dating back to the original effort to expand the typical backpropagation algorithm, BPTT has been formulated to handle the difficulties of temporal sequences that are inherent in sequential data \parencite{werbos1990bptt}. This algorithm allows RNNs in learning sequence dependent data by unfold the network over time steps and then updating weights matrix through the gradient of loss function with respect to the variable \parencite{rumelhart1986backpropagation}.
\\[2ex]
\textbf{Conceptual Framework of BPTT}
\\[1ex]
BPTT works based on the technique of treating an RNN as a deep feedforward network for across multiple time steps. In the forward pass, the RNN, like other artificial neuronal network, applies operation over the data input in sequence, bringing changes in its own state variables at every time step, depending on the input and the previous state of its general working state or hidden state. This sequential processing produces outputs and stores the internal states of the network in any period \parencite{werbos1990bptt}.
\\[1ex]
This unfolds the RNN to construct a traditional Feedforward Neural Network where we can apply backpropagation through time. Below is the conceptual idea of BPTT in RNN.
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.65\textwidth]{../Pic/pic1.png} % Replace "example-image" with your image file
    \caption{Unfolded RNN}
\end{figure}

% Forward pass
\noindent \textbf{Forward Pass}
\\[1ex]
During the forward pass, the RNN processes the input sequence sequentially, computing hidden states and output at each timestep:
\begin{equation}
    h_t = f(U^Tx_{t}+W^Th_{t-1}+b_h)
\end{equation}
\begin{equation}
    \hat{y}_t = f(V^Th_t+b_y)
\end{equation}
\newline  % Computing the loss function
\noindent \textbf{Computing the loss function}
\\[1ex]
Assuming the loss is computed only at the final timestep t:
\begin{equation}
    \mathcal{L}_t = L(y_t, \hat{y}_t)
\end{equation}
In order to do backpropagation through time to tune the parameters in RNN, we need to calculate the partial derivative of loss function $\mathcal{L}$ with respect to the differently parameters.\\
\newline  % Backward pass using the chain rule
\noindent \textbf{Backward pass using the chain rule}
\\[1ex]
Using the chain rule for computing the gradient.\\
Partial derivative of loss function $\mathcal{L}$ with respect to $W$ (hidden to hidden state) at time 2. 
\begin{equation}
    \dfrac{\partial\mathcal{L}_2}{\partial W} = \dfrac{\partial\mathcal{L}_2}{\partial y_2} \cdot \dfrac{\partial y_2}{\partial h_2} \cdot \dfrac{\partial h_2}{\partial h_1} \cdot \dfrac{\partial h_1}{\partial W}
\end{equation}
By mathematic induction 
\begin{equation}
    \dfrac{\partial\mathcal{L}_t}{\partial W} = \dfrac{\partial\mathcal{L}_t}{\partial y_t} \cdot \dfrac{\partial y_t}{\partial h_t} \cdot (\sum_{i=1}^{t} \dfrac{\partial h_t}{\partial h_i} \cdot \dfrac{\partial h_i}{\partial W})
\end{equation}
Where 
\begin{equation}
    \dfrac{\partial h_t}{\partial h_i} = \prod_{j=i+1}^{t} \dfrac{\partial h_j}{\partial h_{j-1}}
\end{equation}

@@

\begin{longtable}{|c|c|c|}
    \hline
    \textbf{Notation} & \textbf{Meaning} & \textbf{Dimension}\\
    \hline
    $U$          & Weight matrix for input to hidden state       & $input\ size\times hidden\ units$\\
    $W$          & Weight matrix for hidden to hidden state      & $hidden\ units\times hidden\ units$\\
    $V$          & Weight matrix for hidden state to output state& $hidden\ units\times number\ of\ class$\\
    $x_t$        & Input vector at time t                        & $input\ size\times 1$\\
    $h_t$        & Hidden state output at time t                 & $hidden\ units\times 1$\\
    $b_h$        & Bias term for hidden state                    & $hidden\ units\times 1$\\
    $b_y$        & Bias term for output state                    & $number\ of\ class\times 1$\\
    $\hat{o}_y$  & Output at time t                              & $number\ of\ class\times 1$\\
    $\hat{y}_t$  & Output at time t                              & $hidden\ units\times 1$\\
    $\mathcal{L}$& Loss at time t                                & $scalar$\\
    \hline
    \caption{Unfolded RNN}
\end{longtable}

@@

\subsubsection{RNN encoder decoder model}
\begin{figure}[!htb]
    \centering
    \includegraphics[width=1\textwidth]{../Pic/RNN_encoder_decoder.jpg} % Replace "example-image" with your image file
\end{figure}
RNN Encoder-Decoderis a RNN based architecture which commonly used for handle sequence to sequence problem, for example machine translation, summarization, and question answering. The designed of this architecture is good for handle the input sequence and the output sequence is not the same length, leveraging the sequential nature of data. 
\\[1ex]
In this architecture, there are two main components, the first component is rnn encoder, the usage of the encoder is to encode the input sequence into a fixed length vector. And the decoder is to decode the vector into the target sequence. 
\\[1ex]
\textbf{Encoder}
\\[1ex]
The encoder is a RNN based block, which processes the input sequence step by step and calculate the hidden state representation for each input token. Let the input sequence be $X = [x_1, x_2, \cdots, x_T]$ where $T$ is the length of the input sequence, and $x_t$ is the input at time step $t$. We use the below formula to computes the hidden states $h_t$.
\\[1ex]
For the Vanilla RNN encoder
\begin{equation}
    h^{enc}_t = tanh(W^{enc}_{ih}x_t + W^{enc}_{hh}h_{t-1}+b^{enc}_h)
\end{equation}

\begin{longtable}{|c|c|c|}
    \hline
    \textbf{Notation} & \textbf{Meaning} & \textbf{Dimension}\\
    \hline
    $W^{enc}_{ih}$          & Weight matrix for input to hidden state in encoder      & $hidden\ units\times input\ size$\\
    $W^{enc}_{hh}$          & Weight matrix for hidden to hidden state in encoder     & $hidden\ units\times hidden\ units$\\
    $b^{enc}_{h}$           & Bias vector for the hidden state in encoder             & $hidden\ units\times hidden\ units$\\
    $h^{enc}_{t}$           & Hidden state at timestep t in encoder                   & $hidden\ units\times hidden\ units$\\
    \hline
    \caption{Encoder}
\end{longtable}
After the model processed the entire input sequence from $t=1$ to $t=T$, the last hidden state $h^{enc}_T$ treat as the context vector $c$. 
\begin{equation}
    c = h^{enc}_T
\end{equation}
\textbf{Decoder}
\\[1ex]
The decoder is the second component of RNN encoder decoder, the usage of the decoder is to generates the output sequence step by step, conditioned on the context vector $c$ and its previous hidden state. At each timestep $t$, the decoder updates its hidden state $s_t'$ based on the previous hidden state $s_{t'-1}$ and the embedding of the previous output $y_{t'-1}$. And we will define the $s_0 = c$ first.
\begin{equation}
    h^{dec}_t = tanh(W^{dec}_{ih}y_{t-1}+W^{dec}_{hh}+b^{dec}_h)
\end{equation}
After the model computed the $h^{dec}_t$, the model will compute the output probability distribution.
\begin{equation}
    P(y_t|y_{t-1}, c) = softmax(W_oh^{dec}_t+b^{dec}_o)
\end{equation}
And the predicted token $y_t = arg maxP(y_t)$, the process will repeat until the decoder predicts the end of sequence token $(<EOS>)$.
\begin{longtable}{|c|c|c|}
    \hline
    \textbf{Notation} & \textbf{Meaning} & \textbf{Dimension}\\
    \hline
    $W^{dec}_{ih}$          & Weight matrix for input to hidden      & $hidden\ units\times input\ size$\\
    $W^{dec}_{hh}$          & Weight matrix for hidden to hidden     & $hidden\ units\times hidden\ units$\\
    $b^{dec}_{h}$           & Bias vector for the hidden              & $hidden\ units\times hidden\ units$\\
    $h^{dec}_{t}$           & Hidden state at timestep t                    & $hidden\ units\times hidden\ units$\\
    $b^{dec}_{o}$           & Bias vector for the output layer             & $hidden\ units\times hidden\ units$\\
    $W^{dec}_{o}$           & Weight matrix for hidden to output     & $output\ feature\ space\times hidden\ units$\\
    \hline
    \caption{Decoder}
\end{longtable}

@@

\textbf{Bahdanau (Additive) Attention}
\\[1ex]
This method proposed by \parencite{bahdanau2016neuralmachinetranslationjointly}, in which this attention uses a feedforward network to compute the attention scores. 
\\[1ex]
\begin{equation}
    e_{t',t} = v_a^Ttanh(W_ah^{dec}_{t'-1}+U_ah^{enc}_t+b_a)
\end{equation}                    
\textbf{Luong (Multiplicative) Attention}
\\[1ex]
This method is introduced by Luong et al \parencite{luong2015effectiveapproachesattentionbasedneural}. This method uses the dot product to compute the attention scores.
\begin{equation}
    e_{t',t} = h^{dec}_{t'-1} \cdot h^{enc}_t
\end{equation}
\textbf{Scaled Dot-Product Attention}
\\[1ex]
The scaled dot-product attention was introduced by \parencite{vaswani2017attentionneed}, this is the key component in Transformer architecture. It improves from the Luong's dot-product attention \parencite{luong2015effectiveapproachesattentionbasedneural} by addressing he numerical stability issues when working with the high dimensional vectors and making attention computation more efficient for parallelization. Below is the formula of attention scores.
\begin{equation}
    S = \dfrac{QK^T}{\sqrt{d_k}}
\end{equation}
Where $S \in \mathbb{R}^{n \times n}$ is pairwise similarity scores between all tokens.
